---
title: "Caregivers and PTSD Project Notebook"
output: html_document
date: "2026-02-09"
---

## Load Packages
```{r packages}
library(tidyverse, quietly = TRUE) # Data wrangling
library(brms)      # Bayesian modeling
library(tidybayes) # For post-processing
library(cmdstanr)  # MCMC sampling
library(lavaan)    # Internal consistency
library(dagitty)   # Specifying & Analyzing DAGs
library(ggdag)     # Drawing DAGs
library(ggtext)    # Fonts for graphs 
library(showtext)  # To display fonts 
library(posterior) # Extract posterior draws

```


## Import Data
```{r import}
source(here::here("src/import.R"))
```


## MCMC Sampling
```{r mcmc}
options(mc.cores = parallel::detectCores()) 
options(brms.backend = "cmdstanr")
```


## Theme Settings
```{r theme}
font_add_google("Karla", family = "karla")
font_add_google("Cardo", family = "cardo")
showtext_auto()
colors_johnson <- MetBrewer::MetPalettes$Johnson[[1]]
c("#a00e00", "#d04e00", "#f6c200", "#0086a8", "#132b69")


theme_density <-
  theme(
    # Remove the legend
    legend.position = 'none',
    
    # Panel customization
    panel.background = element_rect(fill = "white", color = "white"),
    panel.grid.major.x = element_line(color = "#e9e9e9", linewidth = .3),
    panel.grid.minor.x = element_line(color = "#e9e9e9", linewidth = .1),
    panel.grid.major.y = element_blank(),
    
    # Axis customization
    #axis.line = element_line(color = "#3e3e3e"),
    axis.line.y.left = element_blank(),
    axis.title = element_markdown(size = 10, face = "bold"),
    axis.title.y = element_markdown(margin = margin(r = 20)),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    
    # Title customization
    plot.title = element_markdown(size = 14, face = "bold"),
    plot.subtitle = element_markdown(size = 12, color = "#3e3e3e"),
    plot.caption = element_markdown(hjust = 0, color = "#7e7e7e"),
    
    # Strip customization
    strip.background = element_rect(fill = "white"), 
    strip.text = element_text(face = "bold")
  )

```

## Scientific Model

### The Initial Request

#### 1. PTSD and Depression, Anxiety, and Stress
>Depression/anxiety/stress (from DASS-21) levels in Caregivers with high PTSD vs. Caregivers with low PTSD (determined by mean-split).
>
> H1: Depression, Anxiety, and Stress will be significantly higher in caregivers with high probable PTSD than caregivers with low probable PTSD.
>
> - MANOVA
> - IV: PTSD group (High vs. Low)
> - DVs: DASS-21 Depression, Anxiety, Stress subscale scores (continuous)
> - Follow-up with univariate ANOVA for each DASS subscale.


 
#### 2. Burnout and PTSD
> Burnout (from BAT-12) in Caregivers with high PTSD vs. Caregivers with low PTSD(determined by mean-split).
>
> H2: Burnout will be significantly higher in caregivers with high probable PTSD than caregivers with low probable PTSD.
>
> - Independent samples t-test
> - IV: PTSD group (High vs. Low)
> - DV: Burnout score (BAT-12 total)
> - Report: t, p, Cohen’s d, confidence intervals, means


#### Veteran Caregivers and PTSD
> 3. Prevalence of probable PTSD (from PC-PTSD-5) in caregivers to veterans vs caregivers to non-veterans.
>
> H3: Probably PTSD will be more likely in caregivers to veterans than caregivers to non-veterans.
>
> - Chi-square test of independence
> - IV: Caregiver group (veteran caregiver vs. non-veteran caregiver)
> - DV: Probable PTSD status (yes/no)
> - Follow-up with binary logistic regression and include relevant covariates (age, sex, hrs caregiving/week, relationship with care recipient, military status).

#### Veteran Caregivers and Depression, Anxiety, and Stress
> 4. Depression/anxiety/stress levels (from DASS-21) in caregivers to veterans vs caregivers to non-veterans.
>
> H4: Depression, Anxiety, and Stress will be significantly higher in caregivers to veterans than caregivers to non-veterans.
>
> - MANOVA
> - IV: Caregiver group (veteran vs. non-veteran)
> - DVs: DASS-21 Depression, Anxiety, Stress subscale scores (continuous)
> - Follow-up with univariate ANOVA for each DASS subscale.


### My take on this: 
1, 3, and 4 are quite coherent. They concern PTSD, veteran status of care recipient, and depression, anxiety, and stress. 2 regards burnout and is somewhat tangential. Burnout could be combined into the bigger model, but then the question comes: is it downstream from stress/anxiety/depression, or a mediator, or a moderator or what? And the cross-sectional data don't support disentangling this.

* After speaking in person with AN, I have a better idea of her thinking. The idea is that burnout is a proxy for quality of care. Stress, anxiety, and depression matter in their own right for the individual experiecing them, but they also matter to the care recipient if it weakens the quality of care. Absent a measure of quality of care in this data set, burnout is a potential proxy. There are some other potential proxies we could look at, including certain forms of coping (e.g., behavioral disengagement) or caregiver-family conflict. 

The idea, then, would be to model burnout or behavioral disengagement as a downstream effect of DAS. While the data are cross sectional, and therefore give us no clue to the temporal ordering, we can at least think through their association. 

***
The other thing that came out of our conversation was about the role of both veteran status of care recipient and caregiver. Looking at one and not the other could confound the estimate. 

I plan to update the DAG to include both. 

Other potential confounders to model:
- Age, Gender, Sex, Relationship, Income --> all seem reasonable, although we could quibble if income is pre-treatement (e.g., it could be affected by veteran status; so could education...)


***

The MANOVA angle makes sense. Stress, depression, and anxiety can be caused by traumatic stress disorders such as PTSD.

But stress, depression, and anxiety are correlated mental health issues; they tend to hang together. Moreover, people differ in personal traits that can affect these different issues simultaneously.

Given these individual traits, a person might tend to score high on stress, anxiety, and depression relative to what PTSD status alone would predict. So we need to model these 3 outcomes together, all predicted by PTSD, but also allowed for the part that isn't predicted by PTSD (i.e., "error") to co-vary apart from PTSD. 

Now, to consider the veteran variable. It is defensible that having a veteran care recipient might be associated with PTSD because of secondary traumatic stress from a traumatic incident that befell; or because the veteran is more likely to experience trauma after leaving the service, the caregiver might be with them at the time. 

The V->P temporal order is defensible and often assumed. But it too is confounded by childhood experiences that influence both the propensity to join the military and trauma response. Veterans research has not accounted for these pre-service confounders, to my knowledge, likely due to the difficulty of obtaining that data reliably.

### Estimands

Difference in stress, depression, and anxiety between caregivers to veterans and non-veterans.
Difference in PTSD between caregivers to veterans and non-veterans.



## Examine Measures

### Depression DASS-21
```{r omega-depression}
fit_depression <- 
  lavaan::cfa(model = "depression =~ dass_3 + dass_5 + dass_10 + dass_13 + dass_16 + dass_17 + dass_21", 
              data,
              std.lv = TRUE,
              ordered = c('dass_3','dass_5','dass_10',
                          'dass_13','dass_16','dass_17',
                          'dass_21'),
              estimator = "WLSMV")
reliability_depression <- 
  semTools::compRelSEM(fit_depression,
                       obs.var = TRUE,
                       tau.eq = FALSE,
                       ord.scale = TRUE,
                       return.df = TRUE)
reliability_depression
```

### Anxiety DASS-21
```{r omega-anxiety}
str_items <- paste0("DASS", c(1,6,8,11,12,14,18))  

fit_anxiety <- 
  lavaan::cfa(model = "anxiety =~ dass_2 + dass_4 + dass_7 + dass_9 + dass_15 + dass_19 + dass_20", 
              data,
              std.lv = TRUE,
              ordered = c('dass_2','dass_4','dass_7',
                          'dass_9','dass_15','dass_19',
                          'dass_20'),
              estimator = "WLSMV")
reliability_anxiety <- 
  semTools::compRelSEM(fit_anxiety,
                       obs.var = TRUE,
                       tau.eq = FALSE,
                       ord.scale = TRUE,
                       return.df = TRUE)
reliability_anxiety
```

### Stress DASS-21
```{r omega-stress}
fit_stress <- 
  lavaan::cfa(model = "stress =~ dass_1 + dass_6 + dass_8 + dass_11 + dass_12 + dass_14 + dass_18", 
              data,
              std.lv = TRUE,
              ordered = c('dass_1','dass_6','dass_8',
                          'dass_11','dass_12','dass_14',
                          'dass_18'),
              estimator = "WLSMV")
reliability_stress <- 
  semTools::compRelSEM(fit_stress,
                       obs.var = TRUE,
                       tau.eq = FALSE,
                       ord.scale = TRUE,
                       return.df = TRUE)
reliability_stress
```


## Examine Variables
### PTSD
```{r}
data %>% count(care_military, pc_ptsd_positive_screen)
```
### DASS
### Veteran




### Generative Model

With frequentist stats, this would be a MANOVA. A MANOVA is basically a multivariate outcome model with correlated residuals.

Let's fit a multivariate Bayesian regression with correlated residuals across stress, anxiety, and depression. 

This will account for shared variance among the outcomes and yield outcome-specific effects analogous to follow-up ANOVAs.


```{r dag}
dag_1 <- 
  dagify(
    P ~ V,
    D ~ P + eD + V,
    A ~ P + eA + V,
    S ~ P + eS + V,
    eS ~~ eA,
    eD ~~ eS,
    eA ~~ eD,
    exposure = 'P', 
    coords = list(
    x = c(V = -.1, P =  -.1, D =   0, A = 0, S =   0, eD = .05, eA = .05, eS =  .05),
    y = c(V = .05, P = -.05, D =  .1, A = 0, S = -.1, eD = .1, eA =  0, eS = -.1))
)
  
dag_1 %>% 
  tidy_dagitty() %>% 
  mutate(
    label = case_when(
      name == "V" ~ "Veteran", 
      name == "P" ~ "PTSD", 
      name == "D" ~ "Depression", 
      name == "A" ~ "Anxiety", 
      name == "S" ~ "Stress",
      name == "eD" ~ "error",
      name == "eA" ~ "error",
      name == "eS" ~ "error",
      .default = name)
    ) %>% 
  ggplot(aes(x = .data$x, y = .data$y, xend = .data$xend, yend = .data$yend)) +
  #geom_dag_point(color = 'white') +
  geom_dag_text(color = 'black', size = 34, family = "karla", fontface = "plain") +
  geom_dag_text(aes(label = label), size = 12, color = colors_johnson[4], nudge_y = -.0185, nudge_x = -.000005, fontface = "italic", family = "cardo") +
  geom_dag_edges(
    data_directed = dag_1 %>% tidy_dagitty() %>% filter(direction == "->"),
    curvature = .5, 
    start_cap = ggraph::circle(12, 'mm'),
    end_cap = ggraph::circle(12, 'mm'), 
    edge_color = "grey10", 
    edge_width = .45
    ) + 
  theme_dag_blank() +
  lims(x = c(-.11,.11))

ggsave(here::here('output/dag.jpg'), height = 4, width = 6)

```

One important implication of this DAG is that PTSD is positioned as a pipe (i.e., a mediator). If we model PTSD alongside veteran status, then the coefficient on Veteran can only be interpreted as a direct effect, not a total effect. 


#### Likelihood
This means the likelihood of the joint outcome $y_i$ is modeled as a multivariate normal distribution with parameters $\mu$ and $\Sigma$.
 
$$
 \begin{align}
 & y_i \sim MVNormal(\mu_i, \Sigma) \\
 \end{align}
$$
 
 $\mu_i$ is modeled as a vector of comprised of the factor of predictor variables ${x}_i^{\top}$ and the regression coefficients for each outcome $\beta_1, \beta_2, \beta_3$.
 
$$
 \begin{align}
\mu_i =
\begin{pmatrix}
\alpha_s + \beta_{sV} V_i + \beta_{sP} P_i \\
\alpha_s + \beta_{aV} V_i + \beta_{aP} P_i \\
\alpha_s + \beta_{dV} V_i + \beta_{dP} P_i
\end{pmatrix}
\end{align}
$$

$\Sigma$ is a residual covariance matrix

$$
\Sigma =
\begin{pmatrix}
\sigma_1^2 & \rho_{12}\sigma_1\sigma_2 & \rho_{13}\sigma_1\sigma_3 \\
\rho_{12}\sigma_1\sigma_2 & \sigma_2^2 & \rho_{23}\sigma_2\sigma_3 \\
\rho_{13}\sigma_1\sigma_3 & \rho_{23}\sigma_2\sigma_3 & \sigma_3^2
\end{pmatrix}
$$

Where $\rho_{12}, \rho_{23}, \rho_{13}$ are the correlated residuals between outcomes -- the extent to which the variation in the outcomes not explained by the IVs is shared among the DVs -- and $\sigma_1, \sigma_2, \sigma_3$ the standard deviations of the residuals which are the unique, unshared error of each outcome. 


#### Priors

When determining priors, always consider the scale of the variables. 

The predictor variables are binary 0-1: Non-Veteran/Veteran and No PTSD/PTSD. The 0 represents the absence of Veteran status or PTSD.

We will standardize the outcomes to make selecting priors easier. This also makes the coefficient standardized rather than unstandardized (e.g., $b$ to $beta$).
 
The intercept represents the mean for the reference group (0s). The beta represents the difference between groups (0-1). $\text{Normal}(0, .5)$ is reasonable for both. That means I think the predicted outcome of the reference group will be +-1SD of the mean in most cases, and most likely within -+0.5 SD. The same for the difference between group means. 


 

##### Standardize the variables
 
```{r}
data <-
  data %>% 
  mutate(dass_depression_z = as.numeric(scale(dass_depression)),
         dass_stress_z = as.numeric(scale(dass_stress)),
         dass_anxiety_z = as.numeric(scale(dass_anxiety)))

```
 
For the prior on the standard error $\sigma$, we need a positive real number, because standard errors are always positive. By standardizing the outcome variables, the variables have been scaled to have a SD = 1. So the prior should be something make make 1 and the values near it highly plausible. Say a half-student-t(3, 0, 1) or a half-Cauchy(0,1)

For the prior on the residual covariance, I'll use the $\text{LJKcorr}(2)$ which is a standard choice for covariance matrices (McElreath, 2015) and is more skeptical of extreme correlations near 1 or -1 than it is correaltions closer to 0.

$$
\begin{align}
& \beta \sim \text{Normal}(0, .5) \\
& \alpha \sim \text{Normal}(0, .5) \\ 
& \sigma \sim \text{half-Student-t}(3, 0, 1) \\ 
& \Sigma \sim \text{LJKcorr}(2) \\ \\
\end{align}
$$

```{r plot-priors}
tibble(`Cauchy(0,1)` = rcauchy(1000, 1, 1),
       `Student-t(3,0,1)` = rstudent_t(n = 1000, 3, 0, 1)) %>% pivot_longer(everything()) %>%  ggplot(aes(value, fill = name)) + geom_density(alpha = .2) + lims(x = c(-.1, 3)) + labs(y = NULL)
```


### Computer Model

#### Simulation Model
```{r}
library(tibble)
library(MASS)  # mvrnorm

set.seed(1)
n <- 200

# predictor
x1 <- rbinom(n, size = 1, prob = .5)
x2 <- rbinom(n, size = 1, prob = .5)

# regression coefficients (one per outcome)
beta_1 <- c(d = 0.5, s = 0.3, a = 0.2)
beta_2 <- c(d = 0.5, s = 0.3, a = 0.2)

# residual covariance
Sigma <- matrix(
  c(1.0, 0.6, 0.5,
    0.6, 1.0, 0.4,
    0.5, 0.4, 1.0),
  nrow = 3
)

# correlated errors
E <- MASS::mvrnorm(n, mu = c(0, 0, 0), Sigma = Sigma)

# outcomes
data_sim <- 
  tibble(
    pc_ptsd_positive_screen = x1,
    caregivee_military = x2,
    dass_depression_z = beta_1["d"] * x1 + beta_2["d"] * x2 + E[,1],
    dass_stress_z = beta_1["s"] * x1 + beta_2["s"] * x2 + E[,2],
    dass_anxiety_z = beta_1["a"] * x1 + beta_2["a"] * x2 + E[,3]
)


```


#### Set Priors
```{r}
priors <- c(
  # Residual correlation shrinkage
  prior(lkj(2), class = "rescor"),

  # Stress
  prior(normal(0, .5), class = "Intercept", resp = "dassstressz"),
  prior(normal(0, .5), class = "b", resp = "dassstressz"),
  prior(student_t(3, 0, 1), class = "sigma", resp = "dassstressz"),

  # Anxiety
  prior(normal(0, .5), class = "Intercept", resp = "dassanxietyz"),
  prior(normal(0, .5), class = "b", resp = "dassanxietyz"),
  prior(student_t(3, 0, 1), class = "sigma", resp = "dassanxietyz"),

  # Depression
  prior(normal(0, .5), class = "Intercept", resp = "dassdepressionz"),
  prior(normal(0, .5), class = "b", resp = "dassdepressionz"),
  prior(student_t(3, 0, 1), class = "sigma", resp = "dassdepressionz")
)

```




#### Fit the simulation data to the model
```{r}
fit_manova_simulation <- 
  brm(
    bf(dass_stress_z ~ 1 + pc_ptsd_positive_screen + caregivee_military) + 
    bf(dass_anxiety_z ~ 1 + pc_ptsd_positive_screen + caregivee_military) + 
    bf(dass_depression_z ~ 1 + pc_ptsd_positive_screen + caregivee_military) + 
    set_rescor(TRUE),
  prior = priors,
  data = data_sim,
  family = gaussian(),
  chains = 4, cores = 4, warmup = 2000, iter = 6000
)

fit_logistic_simulation <- 
  brm(
    bf(pc_ptsd_positive_screen ~ 1 + caregivee_military, family = bernoulli(link = "logit")),
  prior = c(
    prior(normal(0, .5), class = "Intercept"),
    prior(normal(0, .5), class = "b")
  ),
  data = data_sim,
  family = bernoulli(link = "logit"),
  chains = 4, cores = 4, warmup = 2000, iter = 6000
)

```

```{r}
fit_manova_simulation
fit_logistic_simulation
```

The model did a decent job recovering the parameters of the simulated data. 


Plot the posterior residual covariance from the model of simulated data. 

```{r}
fit_manova_simulation %>% 
  tidy_draws() %>% 
  dplyr::select(contains('rescor')) %>% 
  rename(ro_s_a = rescor__dassstressz__dassanxietyz, 
         ro_s_d = rescor__dassstressz__dassdepressionz,
         ro_a_d = rescor__dassanxietyz__dassdepressionz) %>%
  pivot_longer(everything()) %>% 
  ggplot(aes(value, fill = name)) +
  geom_density(alpha = .5)
  

```

##### Try it with stan
One issue is that brms has problems fitting both models simultaneously. It can't do the residual correlation on only the continuous variables -- it also want to do the logistic variable. 


```{r}
stan_data <- 
  list(
    N = nrow(data_sim),
    #X = as.matrix(data_sim %>% dplyr::select(pc_ptsd_positive_screen, caregivee_military) %>% rename(P = 1, V = 2)),
    #K = c(2), 
    V = data_sim$caregivee_military,
    P = data_sim$pc_ptsd_positive_screen,
    Y = as.matrix(
    data_sim[, c("dass_stress_z",
                 "dass_anxiety_z",
                 "dass_depression_z")])
  )

# Compile the stan code
mod <- cmdstan_model("model.stan")

# Fit the model
fit <- mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 2000,
  iter_sampling = 6000,
  seed = 5000
)


```

#### Check Sampling Diagnoses
```{r}
fit$cmdstan_diagnose()
```



#### Extract posterior draws

```{r}
draws <- fit$draws() %>% as_draws_df() %>% tibble()
draws %>% names()

draws <- draws %>% dplyr::select(!contains('Y_rep') & !contains('p_ptsd'))

# Fix names
replacements <- c("1" = "S", 
                  "2" = "A", 
                  "3" = "D",
                  "\\," = "_",
                  "\\[" = "_",
                  "\\]" = "")
names(draws) <- str_replace_all(names(draws), replacements)

# get rid of self-correlation = 1
draws <- draws %>% dplyr::select(-c(R_A_A, R_S_S, R_D_D))
```

```{r}
draws
```

#### Posterior Means


#### Poterior Predictions
Difference in posterior predictions for 'total effect' under do(V=v). Where PTSD is considered a function of veteran, like in the generative model.

```{r}
# Extract the Sigma from each draw
Sigma_list <- 
  draws %>%
  transmute(
    sSS = Sigma_S_S, sSA = Sigma_S_A, sSD = Sigma_S_D,
    sAS = Sigma_A_S, sAA = Sigma_A_A, sAD = Sigma_A_D,
    sDS = Sigma_D_S, sDA = Sigma_D_A, sDD = Sigma_D_D
  ) %>%
  pmap(function(sSS, sSA, sSD, sAS, sAA, sAD, sDS, sDA, sDD) {
    matrix(c(
      sSS, sSA, sSD,
      sAS, sAA, sAD,
      sDS, sDA, sDD
    ), nrow = 3, byrow = TRUE)
  })
```

```{r}
inv_logit <- function(x) 1 / (1 + exp(-x))

ppred_doV_one_per_draw <- function(draws, Sigma_list, v) {

  outcome_names <- c("stress", "anxiety", "depression")

  # per-draw PTSD probability under do(V=v)
  w <- inv_logit(draws$alpha_P + draws$gamma_V * v)

  # sample PTSD for the new person within each posterior draw
  P_new <- rbinom(n = length(w), size = 1, prob = w)

  # per-draw mean vectors mu(v, P_new)
  mu_mat <- draws %>%
    transmute(
      stress     = alpha_Y_S + beta_V_S * v + beta_P_S * P_new,
      anxiety    = alpha_Y_A + beta_V_A * v + beta_P_A * P_new,
      depression = alpha_Y_D + beta_V_D * v + beta_P_D * P_new
    ) %>%
    as.matrix()

  # simulate Y_new ~ MVN(mu, Sigma) once per draw
  Yrep <- map2_dfr(
    split(mu_mat, row(mu_mat)),
    Sigma_list,
    ~ {
      y <- MASS::mvrnorm(1, mu = .x, Sigma = .y)
      y <- setNames(as.numeric(y), outcome_names)
      tibble::as_tibble_row(y)
    }
  )

  Yrep %>%
    mutate(
      V_do = v,
      P_new = P_new
    )
}


## simulate v1 and v0 and contrast
set.seed(1)

Y_v1 <- ppred_doV_one_per_draw(draws, Sigma_list, v = 1)
Y_v0 <- ppred_doV_one_per_draw(draws, Sigma_list, v = 0)

contrasts_do <- bind_cols(
  Y_v1 %>% dplyr::select(stress, anxiety, depression) %>% rename_with(~ paste0(.x, "_v1")),
  Y_v0 %>% dplyr::select(stress, anxiety, depression) %>% rename_with(~ paste0(.x, "_v0"))
) %>%
  transmute(
    stress     = stress_v1     - stress_v0,
    anxiety    = anxiety_v1    - anxiety_v0,
    depression = depression_v1 - depression_v0
  )
```


```{r}
contrasts_do %>% 
  pivot_longer(everything()) %>% 
  ggplot(aes(x = value)) +
  stat_density(aes(fill = after_stat(x > 0)), alpha = .7, color = 'black', size = .2) +
  scale_fill_manual(values = c("FALSE" = "#d04e00", "TRUE" = "#132b69"),
                    labels = c("Below/Equal 0", "Above 0")) +
  facet_grid(vars(name)) + 
  theme_minimal() +
  labs(title = 'p(D,S,A|do(V))', x = 'Contrast V = 0-1', fill = "Condition") +
  theme_density

```


```{r}
contrasts_do %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  mutate(veteran_greater = ifelse(value >0, 1, 0)) %>% 
  summarize(perc_veteran_greater = mean(veteran_greater))

```




## Write Up

### Model 
To estimate the difference in depression, anxiety, and stress between caregivers to veterans and non-veterans,
we modeled these responses using a Bayesian multivariate multiple regression. Our model, specified mathematically below, predicts a multivariate response vector of depression, anxiety, and stress as a function of veteran status of the care recipient and PTSD status of the caregiver. This approach entails modeling the covariance between the errors of these responses (Bloome & Schrage, 2021). This covariance statistically models the assumption that unique factors of individuals partially explain residual variance in depression, anxiety, and stress even after accounting for PTSD and veteran status. To control for confounding from factors related to demographics, we stratified by [covariates] as indicated by the causal model. The complete mathematical model with priors is specified below. We estimated the posterior distribution using Hamiltonian Monte Carlo implemented in Stan version X. We validated the model using simulated data and assess convergence by inspecting effective sample size, trace plots, and r-hat. The analysis was implemented in R version as well as Stan, with all model code available at [Link].

### Priors
Priors were chosen through prior predictive simulations so that pre-data predictions span the range of scientifically plausible outcomes. 

In the results, we compare prior and posterior distributions to demonstrate the effects of the data. 


### Results
- Warn against the table 2 fallacy, if applicable
- draw densities instead of intervals
- 




```{r}
ggdag::ggdag_collider_triangle()
ggdag_confounder_triangle()
ggdag_m_bias()
ggdag_equivalent_dags(confounder_triangle())
```
