---
title: "Caregivers and PTSD Project Notebook"
output: html document

---

## Load Packages
```{r}
library(tidyverse) # Data wrangling
library(dagitty)   # Specifying & Analyzing DAGs
library(ggdag)     # Drawing DAGs
library(brms)      # Bayesian modeling
library(tidybayes) # For post-processing
library(lavaan)    # Internal consistency
```


## Import Data
```{r import}
source(here::here("src/import.R"))
```


## MCMC Sampling
```{r}
options(mc.cores = parallel::detectCores()) 
options(brms.backend = "cmdstanr")
```


## Theme Settings


## Examine Measures
```{r}

dass_anxiety
dass_depression
```


## PTSD
```{r}
data %>% count(care_military, pc_ptsd_positive_screen)
```

## PTSD and Depression, Anxiety, and Stress
>1. Depression/anxiety/stress (from DASS-21) levels in Caregivers with high PTSD vs. Caregivers with low PTSD (determined by mean-split).
>
> H1: Depression, Anxiety, and Stress will be significantly higher in caregivers with high probable PTSD than caregivers with low probable PTSD.
>
> - MANOVA
> - IV: PTSD group (High vs. Low)
> - DVs: DASS-21 Depression, Anxiety, Stress subscale scores (continuous)
> - Follow-up with univariate ANOVA for each DASS subscale.
 
### Generative Model

Continuous outcomes: Depression, Anxiety, and Stress modeled as normal with mean

A MANOVA is basically a multivariate outcome model with correlated residuals.
 
Let's fit a multivariate Bayesian regression with correlated residuals across stress, anxiety, and depression. 

This will account for shared variance among the outcomes and yield outcome-specific effects analogous to follow-up ANOVAs.

#### Likelihood
This means the likelihood of the joint outcome $y_i$ is modeled as a multivariate normal distribution with parameters $\mu$ and $\Sigma$.
 
$$
 \begin{align}
 & y_i \sim MVNormal(\mu_i, \Sigma) \\
 \end{align}
$$
 
 $\mu_i$ is modeled as a vector of comprised of the factor of predictor variables ${x}_i^{\top}$ and the regression coefficients for each outcome $\beta_1, \beta_2, \beta_3$.
 
$$
 \begin{align}
\mu_i =
\begin{pmatrix}
\mathbf{x}_i^{\top}\boldsymbol{\beta}_1 \\
\mathbf{x}_i^{\top}\boldsymbol{\beta}_2 \\
\mathbf{x}_i^{\top}\boldsymbol{\beta}_3
\end{pmatrix}
\end{align}
$$

$\Sigma$ is a residual covariance matrix

$$
\Sigma =
\begin{pmatrix}
\sigma_1^2 & \rho_{12}\sigma_1\sigma_2 & \rho_{13}\sigma_1\sigma_3 \\
\rho_{12}\sigma_1\sigma_2 & \sigma_2^2 & \rho_{23}\sigma_2\sigma_3 \\
\rho_{13}\sigma_1\sigma_3 & \rho_{23}\sigma_2\sigma_3 & \sigma_3^2
\end{pmatrix}
$$

Where $\rho_{12}, \rho_{23}, \rho_{13}$ are the correlated residuals between outcomes, and $\sigma_1, \sigma_2, \sigma_3$ are the standard deviations of the residuals. 


#### Priors


$$
\begin{align}
& \beta \sim \text{Normal}(0, .5) \\
& \alpha \sim \text{Normal}(0, .5) \\ 
& \sigma \sim \text{half-Cauchy}(0, 1) \\ 
& \Sigma \sim \text{LJKcorr}(2) \\ \\
\end{align}
$$
### Computer Model

#### Simulation Model
```{r}
library(tibble)
library(MASS)  # mvrnorm

set.seed(1)
n <- 100

# predictor
x <- rbinom(n, size = 1, prob = .5) - .5

# regression coefficients (one per outcome)
beta <- c(d = 0.5, s = 0.3, a = 0.2)

# residual covariance
Sigma <- matrix(
  c(1.0, 0.6, 0.5,
    0.6, 1.0, 0.4,
    0.5, 0.4, 1.0),
  nrow = 3
)

# correlated errors
E <- MASS::mvrnorm(n, mu = c(0, 0, 0), Sigma = Sigma)

# outcomes
data_sim <- 
  tibble(
    pc_ptsd_positive_screen = x,
    dass_depression_z = beta["d"] * x + E[,1],
    dass_stress_z = beta["s"] * x + E[,2],
    dass_anxiety_z = beta["a"] * x + E[,3]
)


```


#### Set Priors
```{r}
priors <- c(
  # Residual correlation shrinkage
  prior(lkj(2), class = "rescor"),

  # Stress
  prior(normal(0, .5), class = "Intercept", resp = "dassstressz"),
  prior(normal(0, .5), class = "b", resp = "dassstressz"),
  prior(cauchy(0, 1), class = "sigma", resp = "dassstressz"),

  # Anxiety
  prior(normal(0, .5), class = "Intercept", resp = "dassanxietyz"),
  prior(normal(0, .5), class = "b", resp = "dassanxietyz"),
  prior(cauchy(0, 1), class = "sigma", resp = "dassanxietyz"),

  # Depression
  prior(normal(0, .5), class = "Intercept", resp = "dassdepressionz"),
  prior(normal(0, .5), class = "b", resp = "dassdepressionz"),
  prior(cauchy(0, 1), class = "sigma", resp = "dassdepressionz")
)

```


#### Fit the simulation data to the model
```{r}
fit_manova_simulation <- 
  brm(
    bf(dass_stress_z ~ 1 + pc_ptsd_positive_screen) + 
    bf(dass_anxiety_z ~ 1 + pc_ptsd_positive_screen) + 
    bf(dass_depression_z ~ 1 + pc_ptsd_positive_screen) + 
    set_rescor(TRUE),
  prior = priors,
  data = data_sim,
  family = gaussian(),
  chains = 4, cores = 4, warmup = 2000, iter = 6000
)
```

```{r}
fit_manova_simulation
```

Plot the residual covariance.

```{r}
fit_manova_simulation %>% 
  tidy_draws() %>% 
  dplyr::select(contains('rescor')) %>% 
  rename(ro_s_a = rescor__dassstressz__dassanxietyz, 
         ro_s_d = rescor__dassstressz__dassdepressionz,
         ro_a_d = rescor__dassanxietyz__dassdepressionz) %>%
  pivot_longer(everything()) %>% 
  ggplot(aes(value, fill = name)) +
  geom_density(alpha = .5)
  

```


#### Data Prep

```{r}
data <-
  data %>% 
  mutate(dass_stress_z = as.numeric(scale(dass_stress)),
         dass_anxiety_z = as.numeric(scale(dass_anxiety)),
         dass_depression_z = as.numeric(scale(dass_depression)),
         pc_ptsd_positive_screen_c = pc_ptsd_positive_screen - .5
         )
```



#### Sample from the posterior

```{r manova}
fit_manova <- 
  brm(
    bf(dass_stress_z ~ 1 + pc_ptsd_positive_screen) + 
    bf(dass_anxiety_z ~ 1 + pc_ptsd_positive_screen) + 
    bf(dass_depression_z ~ 1 + pc_ptsd_positive_screen) + 
    set_rescor(TRUE),
  prior = priors,
  data = data,
  family = gaussian(),
  chains = 4, cores = 4, warmup = 2000, iter = 6000
)

```


```{r}
fit_manova
```
 
 
 
 
## Burnout and PTSD
>2. Burnout (from BAT-12) in Caregivers with high PTSD vs. Caregivers with low PTSD(determined by mean-split).
>
>H2: Burnout will be significantly higher in caregivers with high probable PTSD than caregivers with low probable PTSD.
>
> - Independent samples t-test
> - IV: PTSD group (High vs. Low)
> - DV: Burnout score (BAT-12 total)
> - Report: t, p, Cohen’s d, confidence intervals, means

```{r}



```


## Veteran Caregivers and PTSD
> 3. Prevalence of probable PTSD (from PC-PTSD-5) in caregivers to veterans vs caregivers to non-veterans.
>
> H3: Probably PTSD will be more likely in caregivers to veterans than caregivers to non-veterans.
>
> - Chi-square test of independence
> - IV: Caregiver group (veteran caregiver vs. non-veteran caregiver)
> - DV: Probable PTSD status (yes/no)
> - Follow-up with binary logistic regression and include relevant covariates (age, sex, hrs caregiving/week, relationship with care recipient, military status).

```{r}



```


## Vetean Caregivers and Depression, Anxiety, and Stress
> 4. Depression/anxiety/stress levels (from DASS-21) in caregivers to veterans vs caregivers to non-veterans.
>
> H4: Depression, Anxiety, and Stress will be significantly higher in caregivers to veterans than caregivers to non-veterans.
>
> - MANOVA
> - IV: Caregiver group (veteran vs. non-veteran)
> - DVs: DASS-21 Depression, Anxiety, Stress subscale scores (continuous)
> - Follow-up with univariate ANOVA for each DASS subscale.


```{r}



```


